---
title: "MATH 216 Homework 1"
author: "Nina Sonneborn"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr)
library(hflights)
library(knitr)

# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>% 
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")
```





## Admistrative:

Please indicate

* Who you collaborated with: Amanda, Katherine, Kyler
* Roughly how much time you spent on this HW so far: 5 hours
* The URL of the RPubs published URL [here](http://rpubs.com/nsonneborn/hw1).
* What gave you the most trouble: all about the same
* Any comments you have: see comments in code





## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.

Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
# Add boolean variable "over_30" to show whether departure delay was over 30 min

# Add variable "p_delayed" to show proportion of flight delays over 30 min on each date

flights <- flights %>% 
  mutate(over_30 = ifelse(dep_delay > 30 ,TRUE, FALSE)) %>%
  group_by(date) %>%
  mutate(p_delayed = mean(over_30 == TRUE, na.rm=TRUE))

seasonal_delays <-  flights %>%
  ggplot(aes(x=date, y=p_delayed)) +
  geom_point() +
  ggtitle("Seasonality of Departure Delays at IAH in 2011") +
  xlab("Date") +
  ylab("Proportion of departures delayed for over 30 minutes") + 
  geom_smooth()

seasonal_delays
```
*Note:* Before drawing any conclusions, I should investigate whether there is a pattern to
when there are missing values. By removing them, I could be missing something.

It appears in this plot that outliers may be obscuring seasonal trends by adjusting the y scale and making it hard to see typical trends. Perhaps I could bin into months and then do some box plots to better show the trend that I'm trying to see. I don't want to completely get rid of outliers, as they may say something (for example, there may be more delays on holidays or something).


```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
seasonal_delays_box <-  flights %>%
  ggplot(aes(x=as.factor(lubridate::month(date)), y=p_delayed)) +
  geom_boxplot() +
  ggtitle("Seasonality of Departure Delays at IAH in 2011") +
  xlab("Month") + 
  ylab("Proportion of departures delayed for over 30 minutes") +
  geom_hline(aes(yintercept=median(p_delayed, na.rm = TRUE), col="red"), show.legend = FALSE) 
# how do I make a legend for the red line that says median?

seasonal_delays_box 
```

This still may not be the ideal visualization, but I like it better.
It is definitely worth noting all the outliers. Based on the median, which I think is a good variable to look at here because it isn't skewed by outliers, there is an increase in the proportion of delayed flights during June and July, while there is a notable decrease in September, October and November. 



## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?

```{r, echo=FALSE, fig.width=12, fig.height=6}

oldest_planes <-
  full_join(select(planes, plane, year), flights, by = "plane") %>%
  group_by(carrier) %>%
  summarise(mean_yr = mean(year, na.rm=TRUE)) %>%
  arrange(mean_yr) 

oldest_planes %>% knitr::kable()
```

This shows which airline, on average has the oldest planes. 
This analysis, based on mean year the plane was made per airline, shows that MQ and AA had, on average, the oldest planes. However, this could easily be due to a few really old planes. In fact, when isolating MQ, we can see that only 57 of 4,648 entries have values for the plane's year. It may be good to see the distribution for each carrier.

```{r, echo=FALSE, fig.width=12, fig.height=6}
plane_yrs <- full_join(select(planes, plane, year), flights, by = "plane")

plane_yrs_plot <- plane_yrs %>%
  ggplot(aes(x=year)) +
  geom_histogram() +
  facet_wrap(~carrier, scales = "free") +
  ggtitle("Distribution of manufacture year for planes by carrier")
plane_yrs_plot
```


But what would really be effective is if we could see what year planes started getting crammed and then look at proportion of planes from before that year ...

GOOGLE KNOWS ALL!

[USA today on airline seats](http://www.usatoday.com/story/travel/columnist/mcgee/2014/09/24/airplane-reclining-seat-pitch-width/16105491/) gives some input into the evolution of seat pitch and seat width of commercial airplanes over recent years.
While different airlines have different ranges in seat pitch change, seat width took a dive around 1995.
Thus, I'll look at proportion of planes made before '95 on each airline.
```{r, recho=FALSE, fig.width=12, fig.height=6}
flights <- 
  full_join(select(planes, plane, year), flights, by = "plane") %>%
  group_by(carrier) %>%
  mutate(p_old = mean(year <= 1995, na.rm=TRUE))
      # p_old shows the proportion of planes made before 1995 on each airline

p_old <- flights %>%
  group_by(carrier) %>%
  summarise(proportion_old = format(mean(p_old, na.rm=TRUE), digits=4)) %>%
  arrange(desc(proportion_old))
p_old %>% knitr::kable()
```


This analysis shows MQ, AA, US and DL as the carriers with the most planes made before 1995, and thus probably the most spacious economy seating. The second analysis, in my opinion, is more informative for the traveler who is interested in spacious seats. I would have to discount MQ as too many of the values are missing (discussed above). Based on the histograms I would say US, AA, DL and WN are pretty good bets.

## Question 3:

* What states did Southwest Airlines' **flight paths** tend to fly to?
* What states did Southwest Airlines' **flights** tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.

[Carrier codes found here](https://www.census.gov/foreign-trade/reference/codes/aircarrier/acname.txt)
```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
SWA <- full_join(flights, airports, by=c("dest" = "iata")) %>%
  filter(carrier == "WN")

# For flight, I should count each entry (all the rows)
SWA_flights <- SWA %>%
  count(state, sort = TRUE)

SWA_flights %>% knitr::kable()


# For flight path, I should count each distinct flight number 
SWA_flight_paths <- SWA %>%
  group_by(state) %>%
  summarise(count = n_distinct(flight)) %>% # number of distinct flight #'s to each state
  arrange(desc(count))

SWA_flight_paths %>% knitr::kable()
```





## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to/from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
july <- full_join(flights, airports, by = c("dest"="iata"))
july <- full_join(july, states, by="state") %>%
  filter(lubridate::month(date) == 07) %>%
  select(date, flight, dest, state, region, carrier) 

regional_sums <- tally(group_by(july, carrier, region))

regional_proportions <- regional_sums %>%
  group_by(carrier) %>%
  mutate(carrier_sum = sum(n)) %>%
  mutate(proportion = format(n/carrier_sum, digits=3))
regional_proportions %>% knitr::kable()
# this table is good except the demoninator in calculating the proportion includes the "NA" region.. 
# there aren't too many missing values, but still.
NA_vals <- july %>% filter(is.na(region))
NA_vals
# only 3 "states"" are filed as NA: HI, AK, PR. In fact, then, the NA category is worth keeping, but thinking of as a new category (territories)


regional_sums_plot <- regional_sums %>%
  ggplot(aes(x=region, y = n, fill = region)) +
  geom_bar(stat="identity") +
  facet_wrap(~carrier, ncol = 5, scales = "free_y") +
  ggtitle("Regional distribution of flights out of Houston in July of 2011")
regional_sums_plot
```
Note: asking for "flies to/from Houston" but we only have data for destination.
Note: only 3 "states"" are filed as NA: HI, AK, PR. In fact, then, the NA category is worth keeping, but thinking of as a new category (territories).


